Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-08-06 paper/paper.tex
Add citations

To overcome these design limitations, we have developed a novel automatic type system, "cpplab", which binds "MATLAB" code to "C++" header files. This architecture automatically creates objects in "MATLAB" that represent the underlying object-oriented "C++" code, without having to re-specify these objects in "MATLAB". In this paper, we introduce "xolotl", an implementation of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of "MATLAB" code using a hierarchical and intuitive syntax. Since models in the "MATLAB" workspace are automatically linked to models in the "C++" implementation, configuring these objects in "MATLAB" transparently configures the underlying "C++" objects. "xolotl" comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. The GUI permits real-time manipulation of any network parameter and an immediate visualization of the effect of that parameter on arbitrary dynamical features of the model. The ease-of-use of these tools lends them to pedagogical applications and the rapid prototyping of models. This tool aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

@@

The model can be integrated for various amplitudes of injected current to determine its f-I curve, or the relationship between injected current and its firing rate (cite). Figure \ref{fig:figurehh}E shows the f-I curve of this model, obtained by repeated integration of the model.  Finally, the built-in \texttt{show} function can plot activation ($m$) and inactivation ($h$) curves and voltage-dependent timescales of any channel type in the simulator (Figure \ref{fig:figurehh}D-G). These plots reveal that activation kinetics of the "NaV" channels are much faster than that of the "Kd" channels (Figure \ref{fig:figurehh}F), which facilitates the transient depolarization in an action potential. In summary, the simulator allows the user to construct and integrate this model in a few lines of code, and provides a rich visualization of the dynamics of the model.

@@

Voltage clamping is a experimental technique where a amplifier is configured to inject the appropriate amount of current through a electrode to maintain the voltage of a cell at a desired value (cite D\&A). Under this paradigm, the membrane voltage is "clamped" or fixed to a desired value, permitting the study of voltage-dependent ion channels, since the sum of all currents through the population of ion channels in the cell is equal and opposite to the current injected by the clamp (Figure \ref{fig:figureclamp}A). By combining voltage clamp with the use of pharmacological agents to block all channels but the one of interest, the voltage-sensitivity of an ion channel population can be characterized (\cite{coleIonicCurrentMeasurements1960, coleIonsPotentialsNerve1955, hodgkinEffectSodiumIons1949, hodgkinMeasurementCurrentvoltageRelations1952a, hodgkinQuantitativeDescriptionMembrane1952a, turrigianoSelectiveRegulationCurrent1995}).

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_clamp}
	\caption{Simulating a voltage-clamp experiment. (A) The diagram shows a cell with delayed rectifier potassium conductance (from \cite{liuModelNeuronActivitydependent1998}) that is being recorded from in two-electrode voltage clamp. (B) The code snippet shown here sets up a model with a single compartment and a single channel type, and clamps the cell to a constant voltage and integrates it. (C) Voltage steps that the cell is clamped to. (D) Current injected into the cell required to clamp it to the voltages in (C). (E) Asymptotic clamped current $vs.$ clamped voltage for this cell. (F) Accounting for the reversal potential of Potassium ions yields the conductance-voltage curve of this channel type. (G) Normalized conductance-voltage curves, with sigmoid fits with various exponents. (H) An exponent of $n=4$ yields the best fit, allowing for the characterization of the activation function of this channel type. }
	\label{fig:figureclamp}
\end{figure}

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_mechanism}
	\caption{Modeling intracellular mechanisms. (A) A single-compartment neuron model with 8 channel types. Since there is no mechanism for changing intracellular Calcium in this model, the Calcium level stays constant (B), and the cell tonically spikes (C). Intracellular Calcium buffering and influx through voltage gated Calcium channels (VGCCs) can be modeled using a simple differential equation (D). Code snippet shows how this mechanism can be added to the neuron model (E). The cell now bursts periodically, with synchronized oscillations in intracellular Calcium (F-G). (H) Schematic of Calcium-dependent integral control homeostasis (cite O'Leary et al.). In this feedback system, the rates of mRNA synthesis depend on the Calcium level in the cell, which depends on the membrane voltage, which in turn depends on the conductance density of all channel types, which, through translation, is a function of the mRNA abundance. (I) The code snippet shows how these integral controllers are implemented as \texttt{mechanism} objects, and can be added to conductances. (J) On integrating the model, intracellular calcium levels rise and approach the target (red dashed line). This is accompanied by an increase in the conductance densities of all channels being controlled by this homeostatic mechanism (K). The voltage behavior of the cell changes from silence to bursting with truncated spikes to regular bursting.}
	\label{fig:figuremechanism}
\end{figure}

>>>>

To overcome these design limitations, we have developed a novel automatic type system, "cpplab", which binds "MATLAB" code to "C++" header files. This architecture automatically creates objects in "MATLAB" that represent the underlying object-oriented "C++" code, without having to re-specify these objects in "MATLAB". In this paper, we introduce "xolotl", an implementationclose of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of "MATLAB" code using a hierarchical and intuitive syntax. Since models in the "MATLAB" workspace are automatically linked to models in the "C++" implementation, configuring these objects in "MATLAB" transparently configures the underlying "C++" objects. "xolotl" comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. The GUI permits real-time manipulation of any network parameter and an immediate visualization of the effect of that parameter on arbitrary dynamical features of the model. The ease-of-use of these tools lends them to pedagogical applications and the rapid prototyping of models. This tool aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

@@

The model can be integrated for various amplitudes of injected current to determine its f-I curve, or the relationship between injected current and its firing rate (\cite{kisperskyIncreaseSodiumConductance2012}). Figure \ref{fig:figurehh}E shows the f-I curve of this model, obtained by repeated integration of the model.  Finally, the built-in \texttt{show} function can plot activation ($m$) and inactivation ($h$) curves and voltage-dependent timescales of any channel type in the simulator (Figure \ref{fig:figurehh}D-G). These plots reveal that activation kinetics of the "NaV" channels are much faster than that of the "Kd" channels (Figure \ref{fig:figurehh}F), which facilitates the transient depolarization in an action potential. In summary, the simulator allows the user to construct and integrate this model in a few lines of code, and provides a rich visualization of the dynamics of the model.

@@

Voltage clamping is a experimental technique where a amplifier is configured to inject the appropriate amount of current through a electrode to maintain the voltage of a cell at a desired value (\cite{dayanTheoreticalNeuroscience2001}). Under this paradigm, the membrane voltage is "clamped" or fixed to a desired value, permitting the study of voltage-dependent ion channels, since the sum of all currents through the population of ion channels in the cell is equal and opposite to the current injected by the clamp (Figure \ref{fig:figureclamp}A). By combining voltage clamp with the use of pharmacological agents to block all channels but the one of interest, the voltage-sensitivity of an ion channel population can be characterized (\cite{coleIonicCurrentMeasurements1960, coleIonsPotentialsNerve1955, hodgkinEffectSodiumIons1949, hodgkinMeasurementCurrentvoltageRelations1952a, hodgkinQuantitativeDescriptionMembrane1952a, turrigianoSelectiveRegulationCurrent1995}).

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_clamp}
	\caption{Simulating a voltage-clamp experiment. (A) The diagram shows a cell with delayed rectifier potassium conductance (from \cite{liuModelNeuronActivityDependent1998}) that is being recorded from in two-electrode voltage clamp. (B) The code snippet shown here sets up a model with a single compartment and a single channel type, and clamps the cell to a constant voltage and integrates it. (C) Voltage steps that the cell is clamped to. (D) Current injected into the cell required to clamp it to the voltages in (C). (E) Asymptotic clamped current $vs.$ clamped voltage for this cell. (F) Accounting for the reversal potential of Potassium ions yields the conductance-voltage curve of this channel type. (G) Normalized conductance-voltage curves, with sigmoid fits with various exponents. (H) An exponent of $n=4$ yields the best fit, allowing for the characterization of the activation function of this channel type. }
	\label{fig:figureclamp}
\end{figure}

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_mechanism}
	\caption{Modeling intracellular mechanisms. (A) A single-compartment neuron model with 8 channel types. Since there is no mechanism for changing intracellular Calcium in this model, the Calcium level stays constant (B), and the cell tonically spikes (C). Intracellular Calcium buffering and influx through voltage gated Calcium channels (VGCCs) can be modeled using a simple differential equation (D). Code snippet shows how this mechanism can be added to the neuron model (E). The cell now bursts periodically, with synchronized oscillations in intracellular Calcium (F-G). (H) Schematic of Calcium-dependent integral control homeostasis (\cite{olearyCellTypesNetwork2014}). In this feedback system, the rates of mRNA synthesis depend on the Calcium level in the cell, which depends on the membrane voltage, which in turn depends on the conductance density of all channel types, which, through translation, is a function of the mRNA abundance. (I) The code snippet shows how these integral controllers are implemented as \texttt{mechanism} objects, and can be added to conductances. (J) On integrating the model, intracellular calcium levels rise and approach the target (red dashed line). This is accompanied by an increase in the conductance densities of all channels being controlled by this homeostatic mechanism (K). The voltage behavior of the cell changes from silence to bursting with truncated spikes to regular bursting.}
	\label{fig:figuremechanism}
\end{figure}
